An FPGA Implementation of a Long Short-Term Memory Neural Network

### 2016-07-14 paper/paper.tex
Conference Paper Concluded

%% COMMENT THIS FOR SUBMISSION
\author{\IEEEauthorblockN{Jose Pedro Castro Fonseca} \\ 
\IEEEauthorblockA{Faculty of Engineering University of Porto \\
Porto, Portugal}}
\IEEEspecialpapernotice{(Working Version)}

@@

\begin{abstract}
This work proposes a hardware architecture for a Long Short-Term Memory (LSTM) Neural Network, aiming
to outperform software implementations, by exploiting the inherent parallelism in it.  
The main design decisions are presented, along with the proposed network architecture. A description of the main
building blocks of the network is also presented. The network is synthesized for various sizes,
and performance results are presented and analysed. The synthesized network achieves a 195 times speed-up
over a custom-built software network, proving the benefits of parallel computation for this kind of network.
\end{abstract}

\section{Introduction}\label{sec:intro}
\IEEEPARstart{N}{eural} Networks are one of the most commonly used techniques in Deep Learning. This
particular type of network, LSTM, is a recursive network, given that the neuron outputs in a certain 
time step are also fed as the input in the next time step, and that way their memory elements can make
sense of patterns within data sequences, unlike classical recursive neural networks.
These algorithms have been profusely implemented in software, and their practical applications are plentiful. 
However, the benefits of the inherent parallelism offered by a dedicated hardware
platform are not availed, and there are relatively few implementations of Machine Learning algorithms in 
these kind of platforms.

@@

\section{LSTM Neural Networks}\label{sec:lstmnn}
LSTM Networks were originally formulated in~\cite{Hoch97}, but its formulation has been incrementally 
updated in~\cite{Gers00} and~\cite{Gers2000}, and the most current version is the one in~\cite{Graves05}. 
One of the inital proposers of LSTM, Prof. Jürgen Schmidhüber, did a survey on the most common variations 
of the model last year~\cite{Greff15}, which is the reference for this short discussion.

@@

    \item \textbf{Block Input Gate} -- as the name implies, this gate controls the flow of information 
from the input gate to the memory cell. It also receives the input vector and the previous output 
set, producing an output $\mb{z}^{(t)}$. The \textbf{activation function of this gate can be either}, 
but the most common choice is the \textbf{Hyperbolic Tangent}.

    \item \textbf{Forget Gate} -- its role is to control the contents of the Memory Cell, either to 
set or reset them, using the \textit{Hadamard Elementwise} vector multiplication of its output at time 
$t$, $\mb{c}^{(t)}$, with the contents of the memory unit at the previous time step, $\mb{c}^{(t-1)}$. 
The activation function of this gate is \textbf{always sigmoid}, and the resulting signal is $\mb{f}^{(t)}$.

@@

    \item \textbf{Output Activation} -- the output of the Memory Cell goes through this activation 
function that, as the gate activation function, can be any, but the \textit{hyperbolic tangent} 
is the most common choice.

@@

The operation of each set of gates of the layer is given by the following set of equations
\begin{eqnarray}
    \mb{z}^{(t)} & = g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \nonumber\\
    \mb{i}^{(t)} & = \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \nonumber\\
    \mb{f}^{(t)} & = \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \nonumber\\
    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \nonumber\\
    \mb{c}^{(t)} & = \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \nonumber \\
    \mb{y}^{(t)} & = \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) \label{eq:equationsLSTM}
\end{eqnarray}
where $\odot$ is the Hadamard multiplication. The $i$-th element of the previous vectors in bold corresponds 
to the value of the $i$-th neuron in the layer, which is a very convenient and compact representation of the 
whole layer. Furthermore, if the layer has $N$ LSTM neurons and $M$ inputs (i.e. the size of the layer that 
precedes this), we see that the input weight matrices $\mb{W}_*$ have size $N \times M$, and the  recurrent 
weight matrices $\mb{R}_*$ are squre matrices of size $N \times N$, and that the bias weight matrices $\mb{b}_*$ 
and the matrices $\mb{y}^{(t)}$ through $\mb{c}^{(t)}$ have size $N \times 1$, and are thus, in fact, vectors.

@@

The use of LSTM Networks are plentiful: in \textbf{Handwriting Recognition}~\cite{ICDAR09}, where they surpass 
HMM-based models in optical character recognition~\cite{Breuel13}; in \textbf{Speech Recognition}, where, for 
instance, Graves et al.~\cite{Graves13}, in 2013, set a new record on the TIMT Phoneme Recognition Benchmark, 
and also in \textbf{Large Scale Acoustic Modelling of Speech}~\cite{Sak14}. Other uses include 
\textbf{Handwriting Synthesis}~\cite{Graves13_2}, \textbf{Translation}~\cite{Sustkever14}, \textbf{Biomedical Applications} such
as protein homology detection~\cite{Hochreiter07}, \textbf{Music Analysis and Composition}, such as the transcription of piano music 
to MIDI~\cite{Bock12} and automated composition~\cite{Coca13} and improvisation~\cite{Eck02}, and lastly \textbf{Video and Image Analysis} as
in~\cite{Vinyals14},~\cite{Donahue14} and~\cite{Donahue14_2}.

Hitherto, there is but one actual implementation of an LSTM network in hardware, published recently 
(March 2016) by Chang et al.~\cite{Chang15}. A 2 layer LSTM network (two layers in series) was implemented,
with 128 neurons each, which, for processing 1000 samples, yielded an execution time of $0.932 \, \si{\second}$. Assuming 
that both layers are equal, and thus have the same execution time, this yields an approximate execution time of 
$29.13 \, \si{\micro\second}$ per incoming sample (dividing the total execution time by $2 \times 1000$). 
Therefore, this network is able to perform around 34.3 thousands of forward propagations per second. 
Although the network is large, its performance is well below this work, because it does not have
a full level of parallelism when compared to the proposed design of Section~\ref{sec:proparch}, and it
makes use of external memory to store the weights. This last point is understandable, since the authors use
the FPGA core as a co-processor for the main CPU.

\section{Proposed Architecture}\label{sec:proparch}
Before presenting the 
\subsection{Activation Function Calculation}\label{sec:proprarch_af}
In order to evaluate the transcendental activation functions $\sigma(x)$ and $\tanh(x)$, \textbf{Polynomial Approximations} were used, as detailed in~\cite{Muller05}, since evaluating a polynomial does not have high memory usage needs (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the number of multiplications needed is low enough to not pose a restriction both on resources (now DSP slices, and not memory) and in speed (number of number of clock cycles needed to output a result).

The error minimization strategy used to find the optimal polynomial was the \textbf{Least Maximum Approximation}, where the \emph{maximum} error is \emph{minimized}, making use of Remez's Algorithm, which sets a system of $n+2$ linear equations such as~\ref{eq:remezline}

\begin{equation}\label{eq:remezline}
    p(x_i) - f(x_i) = (-1)^{n+1} \epsilon \Leftrightarrow p_0 + p_1 x_i + p_2 x_i^2 + \cdots + p_n x_i^n - f(x_i) = (-1)^{n+1} \epsilon
\end{equation}
where $i \in \left[0, n+1\right]$. Of course, both functions have horizontal assymptotes, which can be used as the value in the edges, but instead of performing the optimization in the single interval in between the chosen points from where the module evaluates the function as the value of the assymptote, I have further split the interval in 4, in order to have lower degree polynomials -- simpler to evaluate -- at a reduced error cost. The algorithm was run using Python, and targeted second degree polynomials for each interval. After that, the Verilog model was described, where the correct coefficients are loaded according to the interval where the incoming operand is located. The evaluation of the polynomial was accomplished using the \textbf{Horner's Rule}, noting that

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth]{figures/nonlin-out.eps}
    \caption[Plot of the output of the Sigmoid Calculator HDL module]{Plot of the output of the Sigmoid Calculator HDL module}
    \label{fig:nonlin-out-sig}
\end{figure}

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth]{figures/nonlin-out-tanh.eps}
    \caption[Plot of the output of the Hyperbolic Tangent HDL module]{Plot of the output of the Hyperbolic Tangent HDL module}
    \label{fig:nonlin-out-tanh}
\end{figure}

@@

The matrix-vector dot product of a matrix $A$ of size $N \times M$ by a vector $x$ of size $M$, if performed in a linear non-parallel way, can be described in terms of Algorithm~\ref{matvec-alg}

\begin{algorithm}
\begin{algorithmic}
\For {$i$ = $1:N$}
    \For {$j$ = $1:M$}
    \State $y_i := y_i + A_{ij} \cdot x_j$
    \EndFor
\EndFor
\end{algorithmic}
\caption{Matrix-vector multiplication of a matrix}
\label{matvec-alg}
\end{algorithm}
This operation has a computational complexity of $O(n^2)$. Each of the $i$-th components of the output vector $y$ can be calculated \textbf{in parallel}, each only requiring the corresponding $i$-th line from the matrix. Follow this approach, matrix-vector multiplication can now be performed in \textbf{linear time}, which is one of the great advantages of custom-tailored hardware solutions.

@@

Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication units within the module, respectively, where I have set $K_G = 4$, for the same matrix and vector sizes as before. Note that in this situation, we would only have 2 multipliers, and the module would be composed of two multiplication units, such as those in Figure~\ref{fig:array-prod}, that work in parallel. They address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}. The dark shaded part of the memory is used by the first multiplier, and the light shaded is used by the other, in parallel for a fixed \verb+rowMux+ -- this signal is produced by the control unit of the module, and essentially operates the left multiplexer and right demultiplexer of Figure~\ref{fig:array-prod}, that allows to choose the proper position of the weight column and to write to the correct output register, respectively. In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluating $y_0$ and $y_4$. After this, \verb+rowMux+ is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$. Therefore, we have the correct result vector in a time proportional to $K_G \cdot M$: a 3 clock cycle overhead is added to the previous estimate, since the memory only outputs the appropriate column in the \emph{next clock cycle}, and that this module is pipelined both at the input and at the output in order to increase the maximum clock frequency.

\begin{figure}
    \centering
    \includegraphics[width=2.5in]{figures/mem-array-prod.eps}
    \caption[A column of the matrix that serves as input to the module]{A column of the matrix that serves as input to the module. The dark shaded part is for the first multiplier, and the light shaded is for the other, in parallel. The rowMux signal addresses the position within each shaded area}.
    \label{fig:mem-arrayprod}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=2.5in]{figures/array-prod.eps}
    \caption[The $i$-th row multiplication unit of the Module]{The $i$-th row multiplication unit of the Module, where rowMux and colSel are internal signals produced by the control unit of the module. The flip-flop accumulates the sum, and the output demux selects the appropriate memory position on where to store this value, within the slot attributed to this multiplication, from $i\cdot K_G$ to $i\cdot K_G + \left[K_G-1\right]$}
    \label{fig:array-prod}
\end{figure}

@@

\subsection{Gate Module}\label{sec:proparch_gate}
The Gate modules are responsible for producing the internal signal vectors for $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$. If we note that, according to~\cite{Greff15}, the removal of peepholes (the signals $\mb{p}_*$) does not compromise significantly the performance of the network~\ref{sec:struct_lstm}, we can ommit them in order to simplify the Gate Module and reduce the usage of DSP slices. This way, a Gate module needs to perform three tasks

@@

\subsection{Network Architecture}\label{sec:proprarch_net}
Equations~\ref{eq:equationsLSTM} suggest that the signals $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$ do not depend on each other -- they operate only on the current input vector $\mb{x}^{(t)}$ and the previous layer output $\mb{y}^{(t-1)}$ -- and therefore can be calculated in parallel. This way we need four Gate Modules working in parallel, each one with its respective weigth RAMs for $\mb{W}_*$ and $\mb{R}_*$, and followed by the respective activation function calculator (detailed in Section~\ref{sec:nonlincalc}). There are three elementwise multiplications, two for producing signal $\mb{c}^{(t)}$ (which can be done in parallel and then summed elementwise) and one for $\mb{y}^{(t)}$ (which can be done only after applying the activation function $\mb{c}^{(t)}$).

Furthermore, we can avoid a naïve translation of the Equations~\ref{eq:equationsLSTM}, which would replicate unnecessary resources (such as elementwise multipliers and activation function calculatores) and require a higher area usage to save a negligible amount of clock cycles, by noting that one of the operands is the output of a $\tanh(\mb{x})$ block and the other of a $\sigma(\mb{x})$, and they are then multiplied (elementwise, of course) together. This way, instead of replicating these $\tanh$-$\sigma$-$(\cdot)$wise structures, we use a \emph{single one} and choose the operand accordingly to the state that the network is currently in. The issue about the elementwise multiplier for $\mb{c}^{(t)}$, which does not use the $\tanh$ activation function, can be solved by adding another multiplexer that chooses between the output of the $\tanh(\mb{x})$ module or the signal $\mb{c}^{(t-1)}$. These ideas resulted in the LSTM network design of Figure~\ref{fig:network-opt}, which is mathematically equivalent to Equations~\ref{eq:equationsLSTM}.

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{figures/network-opt.eps}
    \caption[Block diagram of the hardware LSTM Neural Network]{Block diagram of the hardware LSTM Neural Network}
    \label{fig:network-opt}
\end{figure}
The two left multiplexers control the operands that are fed to the activation function modules, and the selecting signal is generated by the network's state machine, and its value is incremented after each complete usage of the $\tanh$-$\sigma$-$(\cdot)$wise structure: this is where the time multiplexing of the structure takes place. Since in state $\verb+Sel+=1$ the left operand of the elementwise multiplier (the one that preceded the flip-flops in the previous design) is the signal $\mb{c}^{(t-1)}$, another multiplexer was added before the elementwise multiplication, to select the $\mb{c}^{(t-1)}$ signal in that particular case, and the output from the $\tanh(\mb{x})$ block, otherwise.

@@

\begin{equation}\label{eq:numcc_network-opt}
    (N \cdot K_G + 6) + 27  = 33 + N\cdot K_G
\end{equation}
which is only 13 clock cycles more than a fully-parallel, naïve architecture. For instance, an $N=32$ neuron network would require 320 DSP slices, while this new architecture would only require 160, only at the expense of 13 more clock cycles.

@@

\subsection{Validation}

\subsection{Synthesis}

\subsection{Performance}

\section{Conclusion}\label{sec:concl}

>>>>

%% COMMENT THIS FOR SUBMISSION
%%\IEEEauthorblockA{Faculty of Engineering University of Porto \\
%%Porto, Portugal}}
%%\IEEEspecialpapernotice{(Working Version)}

@@

\begin{abstract}
This work proposes a hardware architecture for a Long Short-Term Memory (LSTM) Neural Network, aiming
to outperform software implementations, by exploiting its inherent parallelism.
The main design decisions are presented, along with the proposed network architecture. A description of the main
building blocks of the network is also presented. The network is synthesized for various sizes and platforms,
and the performance results are presented and analysed. The synthesized network achieves a 251 times speed-up
over a custom-built software network, proving the benefits of parallel computation for this kind of network.
\end{abstract}

\section{Introduction}\label{sec:intro}
\IEEEPARstart{N}{eural} Networks are one of the most commonly used techniques in Deep Learning. This
particular type of network, LSTM, is a recursive network, given that the neuron outputs in a certain
time step are also fed as the input in the next time step, and that their memory elements can make
sense of patterns within data sequences, unlike classical recursive neural networks.
These algorithms have been profusely implemented in software, and their practical applications are plentiful.
However, the benefits of the inherent parallelism offered by a dedicated hardware
platform are not availed, and there are relatively few implementations of Machine Learning algorithms in
these kind of platforms.

@@

\section{LSTM Neural Networks}\label{sec:lstmnn}
LSTM Networks were originally formulated in~\cite{Hoch97}, but its formulation has been incrementally
updated in~\cite{Gers00} and~\cite{Gers2000}, and the most recent version is~\cite{Graves05}.
One of the inital proposers of LSTM, Prof. Jurgen Schmidhuber, did a survey on the most common variations
of the model last year~\cite{Greff15}, which is the reference for this short discussion.

@@

    \item \textbf{Block Input Gate} -- as the name implies, this gate controls the flow of information
from the input gate to the memory cell. It also receives the input vector and the previous output
set, producing an output $\mb{z}^{(t)}$. The \textbf{activation function of this gate can be either},
but the most common choice is the \textbf{hyperbolic hangent}.

    \item \textbf{Forget Gate} -- its role is to control the contents of the Memory Cell, either to
set or reset them, using the \textit{Hadamard Elementwise} vector multiplication of its output at time
$t$, $\mb{c}^{(t-1)}$. The activation function of this gate is \textbf{always sigmoid}, and the resulting
signal is $\mb{f}^{(t)}$.

@@

    \item \textbf{Output Activation} -- the output of the Memory Cell goes through this activation
function that, as the gate activation function, can be any, but the \textbf{hyperbolic tangent}
is the most common choice.

@@

The operation of each set of gates of the layer is given by the following set of equations
\begin{eqnarray}
    \mb{z}^{(t)} & = & g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \nonumber\\
    \mb{i}^{(t)} & = & \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \nonumber\\
    \mb{f}^{(t)} & = & \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \nonumber\\
    \mb{o}^{(t)} & = & \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \nonumber\\
    \mb{c}^{(t)} & = & \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \nonumber \\
    \mb{y}^{(t)} & = & \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) \label{eq:equationsLSTM}
\end{eqnarray}
where $\odot$ is the Hadamard multiplication. The $i$-th element of the previous vectors in bold corresponds
to the value of the $i$-th neuron in the layer, which is a very convenient and compact representation of the
whole layer. Furthermore, if the layer has $N$ LSTM neurons and $M$ inputs (i.e. the size of the layer that
precedes this), we see that the input weight matrices $\mb{W}_*$ have size $N \times M$, and the  recurrent
weight matrices $\mb{R}_*$ are square matrices of size $N \times N$, and that the bias weight matrices $\mb{b}_*$
and the matrices $\mb{y}^{(t)}$ through $\mb{c}^{(t)}$ have size $N \times 1$, and are thus, in fact, vectors.

@@

The uses of LSTM Networks are plentiful: in \textbf{Handwriting Recognition}~\cite{ICDAR09}, where they surpass
HMM-based models in optical character recognition~\cite{Breuel13}; in \textbf{Speech Recognition}, where, for
instance, Graves et al.~\cite{Graves13}, in 2013, set a new record on the TIMT Phoneme Recognition Benchmark,
and also in \textbf{Large Scale Acoustic Modelling of Speech}~\cite{Sak14}. Other uses include
\textbf{Handwriting Synthesis}~\cite{Graves13_2}, \textbf{Translation}~\cite{Sustkever14}, \textbf{Biomedical Applications} such
as protein homology detection~\cite{Hochreiter07}, \textbf{Music Analysis and Composition}, such as the transcription of piano music
to MIDI~\cite{Bock12} and automated composition~\cite{Coca13} and improvisation~\cite{Eck02}, and lastly \textbf{Video and Image Analysis} as
in~\cite{Vinyals14},~\cite{Donahue14} and~\cite{Donahue14_2}.

Hitherto, there is but one actual implementation of an LSTM network in hardware, published recently
(March 2016) by Chang et al.~\cite{Chang15}. A 2 layer LSTM network (two layers in series) was implemented,
with 128 neurons each, which, for processing 1000 samples, yielded an execution time of $0.932 \, \si{\second}$. Assuming
that both layers are equal, and thus have the same execution time, this yields an approximate execution time of
$29.13 \, \si{\micro\second}$ per incoming sample (dividing the total execution time by $2 \times 1000$).
Therefore, this network is able to perform around 34.3 thousands of forward propagations per second.
Therefore ts performance is well below this work, which sits at \textbf{0.487 millions} of forward propagations per second.
This is because the work in~\cite{Chang15} does not have a full level of parallelism when compared to the proposed design of
Section~\ref{sec:proparch}, and it makes use of external memory to store the weights. This last point is understandable,
since the authors use the FPGA core as a co-processor for the main CPU, and not as a standalone implementation.

\section{Proposed Architecture}\label{sec:proparch}

The building blocks of the network in Section~\ref{sec:proprarch_net} are presented in Sections~\ref{sec:proprarch_af} through~\ref{sec:proparch_gate}.
The number representation system used for this network was a \textbf{signed fixed-point} Q6.11 system in two's complement. The are 7 integer bits (one of which is
the sign bit), and 11 fractional bits, adding to a total of \textit{18 bits}. The reason we use 18 bits, is to make full use of the DSP48E1 slices available
in the FPGAs.

\subsection{Activation Function Calculation}\label{sec:proprarch_af}
In order to evaluate the transcendental activation functions $\sigma(x)$ and $\tanh(x)$, \textbf{Polynomial Approximations} were used, as detailed in~\cite{Muller05},
since evaluating a polynomial does not have high memory usage needs (as opposed to Table Methods, for instance) and, if the polynomial degree is sufficiently low, the
number of multiplications needed is low enough to not pose a restriction both on resources (now DSP slices, and not memory) and in speed (number of number of clock cycles
needed to output a result).

The error minimization strategy used to find the optimal polynomial was the \textbf{Least Maximum Approximation}, where
the \emph{maximum} error is \emph{minimized}, making use of Remez's Algorithm, which sets a system of $n+2$ linear equations such as Equation~\ref{eq:remezline}

\begin{eqnarray}\label{eq:remezline}
    & p(x_i) - f(x_i) = (-1)^{n+1} \epsilon \Leftrightarrow  \nonumber \\
		& p_0 + p_1 x_i + p_2 x_i^2 + \cdots + p_n x_i^n - f(x_i) = (-1)^{n+1} \epsilon
\end{eqnarray}
where $i \in \left[0, n+1\right]$. Of course, both functions have horizontal assymptotes, which can be used as the value in the edges,
but instead of performing the optimization in the single interval in between the chosen points from where the module evaluates the function
as the value of the assymptote, the interval was further split in 4, in order to have lower degree polynomials -- simpler to evaluate -- at a reduced error cost.
The algorithm was run using Python, and targeted second degree polynomials for each interval. After that, the Verilog model was described, where the correct
coefficients are loaded according to the interval where the incoming operand is located. The evaluation of the polynomial
was accomplished using the \textbf{Horner's Rule}, noting that

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth]{figures/nonlin-out.eps}
    \caption{Plot of the output of the Sigmoid Calculator HDL module}
    \label{fig:nonlin-out-sig}
\end{figure}

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth]{figures/nonlin-out-tanh.eps}
    \caption{Plot of the output of the Hyperbolic Tangent HDL module}
    \label{fig:nonlin-out-tanh}
\end{figure}

@@

The matrix-vector dot product of a matrix $A$ of size $N \times M$ by a vector $x$ of size $M$, if performed in a linear non-parallel way, can be described in terms of Algorithm~\ref{alg:matvec-alg}

\begin{algorithm}
\begin{algorithmic}
\For {$i$ = $1:N$}
    \For {$j$ = $1:M$}
    \State $y_i := y_i + A_{ij} \cdot x_j$
    \EndFor
\EndFor
\end{algorithmic}
\caption{Matrix-vector multiplication of a matrix}
\label{alg:matvec-alg}
\end{algorithm}
This operation has a computational complexity of $O(n^2)$. Each of the $i$-th components of the output vector $y$ can be calculated \textbf{in parallel},
each only requiring the corresponding $i$-th line from the matrix. Following this approach, matrix-vector multiplication can now be performed in \textbf{linear time},
which is one of the great advantages of custom-tailored hardware solutions.

@@

Figures-\ref{fig:mem-arrayprod} and~\ref{fig:array-prod} depict a diagram of the memory access for the Matrix, and the row multiplication
units within the module, respectively, where $K_G = 4$ and for the same matrix and vector sizes as before. Note that in this situation,
we would only have 2 multipliers, and the module would be composed of two multiplication units, such as those in Figure~\ref{fig:array-prod},
that work in parallel. They address a particular column using the signal \verb+colSel+, which is used by the RAM module to output the corresponding
column of the matrix (in regard to the input vector, obviously this signal selects only a single position), depicted in Figure~\ref{fig:mem-arrayprod}.
The dark shaded part of the memory is used by the first multiplier, and the light shaded is used by the other, in parallel for a
fixed \verb+rowMux+ -- this signal is produced by the control unit of the module, and essentially operates the left multiplexer and right demultiplexer
of Figure~\ref{fig:array-prod}, that allows to choose the proper position of the weight column and to write to the correct output vector position, respectively.
In this example, for \verb+rowMux+=0, the control unit increments \verb+colSel+ from 0 to $M$, and thus evaluating $y_0$ and $y_4$. After this, \verb+rowMux+
is incremented to 1, and the process repeats until \verb+rowMux+ reaches $K_G-1$. Therefore, we have the correct result vector in a time
proportional to $K_G \cdot M$: a 3 clock cycle overhead is added to the previous estimate, since the memory only outputs the appropriate column
in the \emph{next clock cycle}, and that this module is pipelined both at the input and at the output in order to increase the maximum clock frequency.

\begin{figure}
    \centering
    \includegraphics[width=2.5in]{figures/mem-array-prod.eps}
    \caption{A column of the matrix that serves as input to the module. The dark shaded part is for the first multiplier, and the light shaded is for the other, in parallel. The rowMux signal addresses the position within each shaded area}.
    \label{fig:mem-arrayprod}
\end{figure}

\begin{figure*}[!t]
    \centering
    \includegraphics[width=\linewidth]{figures/array-prod.eps}
    \caption{The $i$-th row multiplication unit of the Module, where rowMux and colSel are internal signals produced by the control unit of the module. The flip-flop accumulates the sum, and the output demux selects the appropriate memory position on where to store this value, within the slot attributed to this multiplication, from $i\cdot K_G$ to $i\cdot K_G + \left[K_G-1\right]$}
    \label{fig:array-prod}
\end{figure*}

@@

\subsection{Gate Module}\label{sec:proparch_gate}
The Gate modules are responsible for producing the internal signal vectors for $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$. If we
note that, according to~\cite{Greff15}, the removal of peepholes (the signals $\mb{p}_*$) does not compromise significantly the performance of the network,
we can omit them in order to simplify the Gate Module and reduce the usage of DSP slices. This way, a Gate module needs to perform three tasks

@@

\subsection{Network Architecture}\label{sec:proprarch_net}
Equations~\ref{eq:equationsLSTM} suggest that the signals $\mb{z}^{(t)}$, $\mb{i}^{(t)}$, $\mb{f}^{(t)}$ and $\mb{o}^{(t)}$ do not depend on each other -- they
operate only on the current input vector $\mb{x}^{(t)}$ and the previous layer output $\mb{y}^{(t-1)}$ -- and therefore can be calculated in parallel. This way we
need four Gate Modules working in parallel, each one with its respective weight RAMs for $\mb{W}_*$ and $\mb{R}_*$, and followed by the respective activation
function calculator (detailed in Section~\ref{sec:proprarch_af}). There are three elementwise multiplications, two for producing signal $\mb{c}^{(t)}$ (which
can be done in parallel and then summed elementwise) and one for $\mb{y}^{(t)}$ (which can be done only after applying the activation function $\mb{c}^{(t)}$).

Furthermore, we can avoid a naive translation of the Equations~\ref{eq:equationsLSTM}, which would replicate unnecessary resources (such as elementwise multipliers
and activation function calculatores) and require a higher area usage to save a negligible amount of clock cycles, by noting that one of the operands is the output
of a $\tanh(\mb{x})$ block and the other of a $\sigma(\mb{x})$, and they are then multiplied (elementwise, of course) together. This way, instead of replicating
these $\tanh$-$\sigma$-$(\cdot)$wise structures, we use a \emph{single one} and choose the operand accordingly to the state that the network is currently in. The
issue about the elementwise multiplier for $\mb{c}^{(t)}$, which does not use the $\tanh$ activation function, can be solved by adding another multiplexer that chooses
between the output of the $\tanh(\mb{x})$ module or the signal $\mb{c}^{(t-1)}$. These ideas resulted in the LSTM network design of Figure~\ref{fig:network-opt}, which
is mathematically equivalent to Equations~\ref{eq:equationsLSTM}.

\begin{figure*}
    \centering
    \includegraphics[width=0.9\linewidth]{figures/network-opt.eps}
    \caption[Block diagram of the hardware LSTM Neural Network]{Block diagram of the hardware LSTM Neural Network}
    \label{fig:network-opt}
\end{figure*}
The two left multiplexers control the operands that are fed to the activation function modules, and the selecting signal is generated by the network's state machine,
and its value is incremented after each complete usage of the $\tanh$-$\sigma$-$(\cdot)$wise structure: this is where the time multiplexing of the structure takes place.
Since in state $\verb+Sel+=1$ the left operand of the elementwise multiplier (the one that preceded the flip-flops in the previous design) is the signal $\mb{c}^{(t-1)}$,
another multiplexer was added before the elementwise multiplication, to select the $\mb{c}^{(t-1)}$ signal in that particular case, and the output from the $\tanh(\mb{x})$ block, otherwise.

@@

\begin{equation}\label{eq:numcc_network-opt}
    (N \cdot K_G + 6) + 27  = 33 + N\cdot K_G
\end{equation}
which is only 13 clock cycles more than a fully-parallel, naive architecture. For instance, an $N=32$ neuron network would
require 320 DSP slices, while this new architecture would only require 160, only at the expense of 13 more clock cycles.

@@

\subsection{Validation}\label{sec:res-val}

The functionality of the network was verified against a Python model of an LSTM network that was developed as a reference,
both for the forward propagation of the network, as well as for the training algorithm. Python and Numpy were used rather than MATLAB,
since the former has higher performance at the same level of code complexity. The model was trained using the SPSA method~\cite{Spall98}.

The learning problem presented to both the software and hardware network is the \textbf{addition} of two binary numbers of 8 bits. The $i$-th
bit of each number is fed to the network as a vector, and the network outputs its prediction of the correct value of the $i$-th bit
of the \emph{result}. After the whole number is processed, the memory cells of the LSTM network are reset and a new addition
task can be presented to the network. Even though this seems a rather simple problem, it accounts for all the essential issues at which this
network excels: first, this is a \emph{classification} problem in which the Machine Learning algorithm needs to output a prediction based on
the input feature vector, and that prediction has to take into account the \emph{history} of predictions and inputs so far, because the current
bit is the sum of not only the bits of the two operands, but also the \emph{carry} generated at the last few positions -- this is where the
memory cells of this special Recursive Neural Network come into effect.

After the software network was trained, the trained weights were loaded into the network and several addition problems
were fed to the Network in the testbench, which yielded no errors.

\subsection{Synthesis}\label{sec:results_synth}
The proposed network was first synthesized aiming a Xilinx XC7Z020 SoC, for sizes $N$ of 4, 8, 16 and 32, varying the resource sharing parameter $K_G$, and the number of inputs set
to $M=2$. For a network size of 32 and $K_G=8$, the LUT usage exceeded the LUT resources available in the FPGA, so only lower values of $K_G$ were
successfully synthesized. This is because of the complexity of the input multiplexer of the Matrix-vector Dot Product Module in~\ref{sec:proprarch_dot}. To synthesize the
design for sizes $N$ of 64, 128 and 256, a Virtex-7 VC707 board was used.

\subsubsection{Maximum Frequency}\label{sec:results_synth_maxfreq}
The maximum clock frequency that does not cause timing violations of any sort is plotted in Figure~\ref{fig:maxfreq}. For $N=4$, since there are
only 4 rows to be multiplied, the maximum value of $K_G$ is 4, and hence no synthesis was performed for $K_G = 8$ (N.A.); also,
since $K_G = 2$ and $N=32$, a network such as theses would use $32(8/2+3) = 224$ DSP slices, that exceeds the 220 slices
available in the XC7Z020, so there is no synthesis data for that value (N.D.), as well.

It is clear that with increasing $K_G$, the maximum clock speed decreases, and that decrease is steeper for larger values
of $K_G$. This means that there is a critical path in the Matrix-Vector multiplication unit, whose multiplexer becomes
increasingly complex for higher values of $K_G$. On the other hand, when $K_G$ is the same, smaller networks are faster
than larger networks. The fastest design is an $N=4$ and $K_G = 2$ network, with a clock frequency of $158.228$ MHz, and
the slowest one is an $N=32$ and $K_G=4$ network, clocked at $101.523$ MHz. The reference design used for validation in
Section~\ref{sec:res-val} is an $N=8$ and $K_G=2$ network is clocked at $154.321$ MHz, which yields a clock period of
$6.48$ ns.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/maxfreq.eps}
    \caption{The maximum achievable clock frequency for feasible network sizes in the XC7Z020}
    \label{fig:maxfreq}
\end{figure}

As previously stated, a synthesis targeting a Virtex-7 VC707 platform was performed, in order to synthesize networks of
larger size that, due to the LUT usage restriction, could not be synthesized in the XC7Z020. The maximum clock achievable
for each network size is reported in Table~\ref{tab:maxfreq-virtx7}

\begin{table}
	\caption{Maximum clock frequencies for the VC707}
	\label{tab:maxfreq-virtx7}
    \centering
  \begin{tabular}{ | l | c | }
    \hline
     & $Frequency$  \\
    \hline
    $N=64$ & 140.854 MHz \\
    \hline
    $N=128$ & 140.854 MHz\\
    \hline
    $N=256$ &  ??? MHz \\
\hline
  \end{tabular}

\end{table}
